---
title: "TODO"
author: Astrid Deschênes, Daniel King and David A. Tuveson
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    fig_width: 5
    number_sections: true
pkgdown:
  asis: true
bibliography: BasalClassicalHeatmap.bibtex
vignette: >
  %\VignetteIndexEntry{TODO}
  %\VignettePackage{BasalClassicalHeatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis', warning=FALSE, message=FALSE}
BiocStyle::markdown()

suppressPackageStartupMessages({
  library(knitr)
  library(ComplexHeatmap)
  library(BasalClassicalHeatmap)
})

set.seed(121)
```


<br />

**Package**: `r Rpackage("BasalClassicalHeatmap")`<br />
**Authors**: `r packageDescription("BasalClassicalHeatmap")[["Author"]]`<br />
**Version**: `r packageDescription("BasalClassicalHeatmap")$Version`<br />
**Compiled date**: `r Sys.Date()`<br />
**License**: `r packageDescription("BasalClassicalHeatmap")[["License"]]`<br />


# Licensing 


The `r Githubpkg("adeschen/BasalClassicalHeatmap")` package and the underlying 
`r Githubpkg("adeschen/BasalClassicalHeatmap")` code are distributed under the 
Artistic license 2.0. You are free to use and redistribute this software. 

<br>
<br>


# Citing

If you use this package for a publication, we would ask you 
to cite one of the following.

TODO

<br>
<br>

# Introduction

Pancreatic ductal adenocarcinoma (PDAC) is an aggressive form of cancer. 
Bulk transcriptomic analyses distinguished between two main classes 
of tumors: _classical_ and _basal-like_ . The 
_basal-like_ tumors show a more aggressive clinical behavior [@Moffitt2015].

This package enables Basal-like/Classical classification of 
pancreatic tumor samples by generating a heatmap based on transcriptomic data. 
There is three different lists of genes available in the package: the original 
list from Moffitt et al. 2015 [@Moffitt2015]; the updated list 
from Miyabayashi et al. 2020 [@Miyabayashi2020] and the list from 
Tiriac et al. 2018 [@Tiriac2018] that has been developed for pancreatic ductal 
adenocarcinoma (PDAC) organoids.

<br>
<br>

# Basal-like/Classical classification lists


The available lists are from:

- Moffitt et al 2015
- Tiriac et al 2018
- Miyabayashi et al 2020


The Moffitt et al 2015 corresponds to the original Basal-like/Classical 
classification gene list [@Moffitt2015]. The list contains 50 genes: 
25 Basal-like and 25 Classical genes.

The Tiriac et al 2018 list has been developed to classify PDAC 
organoids [@Tiriac2018]. This list contains 88 genes: 22 Basal-like and 
62 Classical genes.

The Miyabayashi et al 2020 is an update of the original Moffitt 
Basal-like/Classical classification gene list [@Miyabayashi2020]. Gene symbols 
that were not found anymore in most recent versions of Ensembl have been
replaced with their newest symbols. As the Moffitt gene list, this list 
contains 25 Basal-like and 25 Classical genes.

The three gene lists are available in the package:

```{r classificationDataset, eval=TRUE}
## Moffitt et al 2015
data("Moffitt_2015")

## Tiriac et al 2018
data("Tiriac_2018")

## Miyabayashi et al 2020
data("Miyabayashi_2020")

## Each dataset contains 2 columns
## The first column has the gene symbols
## The second column the class assignation (Basal-like or Classical)
head(Moffitt_2015)

```

# Reproducible research

To ensure reproducible results, *set.seed()* function should be call 
before *createHeatmap()*. 


```{r demoSeed, eval=TRUE}
## First, fix the seed value
set.seed(144)

## Load Miyabayashi et al 2020 gene list
data("Miyabayashi_2020")

## Load demo PDAC TCGA expression data
data("demo_PDAC_TCGA")

## Create heatmap
createHeatmap(gene_list = Miyabayashi_2020, rna_data = demo_PDAC_TCGA, 
    row_split=2, column_split=2)

```

<br>
<br>


# Acknowledgments

This work was supported by the Lustgarten Foundation, where David A. Tuveson 
is a distinguished scholar and Director of the Lustgarten Foundation–designated 
Laboratory of Pancreatic Cancer Research. 

<br>
<br>


# Session info

Here is the output of `sessionInfo()` on the system on which this document was 
compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

<br>
<br>

# References

